A Plasma Torus Around a Young Low-Mass Star

### 2025-03-19 paper/ms.tex
update

\begin{abstract}
\normalfont
Roughly one percent of red dwarfs younger than 100 million years
show structured, periodic optical light curves suggestive of
transiting opaque material that corotates with the star
\cite{Rebull2016,Stauffer2017,Rebull2018,Bouma2024}.  The
composition, origin, and even the existence of this material are
uncertain. The main alternative hypothesis is that these complex
periodic variables (CPVs) are explained by complex distributions of
bright or dark regions on the stellar surfaces
\cite{Koen2021}.  Here, we present time-series spectroscopy and
photometry of a rapidly-rotating ($P$=3.9\,hr)
CPV, TIC~141146667. The spectra show coherent sinusoidal Balmer
emission at twice to four times the star's equatorial velocity,
demonstrating the existence of extended clumps of circumstellar
plasma --- a plasma torus.  Given that long-lived condensations of
cool ($10^4$ K) plasma can persist in the hot ($10^6$ K) coronae of
stars with a wide range of masses
\cite{CollierCameron1989,Townsend2005,Dunstone2006,Petit2013,Waugh2022,Daley-Yates2024},
these data support the idea that around the lowest-mass stars, such
condensations can be optically thick; the exact origin of the
material and its microphysical opacity remain unclear.
\end{abstract}

@@

The material's composition -- either pure plasma, or else a dusty plasma
-- can be clarified by time-series optical and infrared
spectrophotometry.  While observations of CPVs in the optical have
previously suggested chromaticity consistent with dust
\cite{Tanimoto2020,Gunther2022,Koen2023}, a gray opacity source such as
electron scattering in a plasma transiting over starspots might also
produce chromatic features \cite{Rackham2018}.  The composition and size
distribution of any dust that is present could be determined
by measuring the extinction curve for one or more CPVs from
1-10\,$\mu$m.  Composition and size distributions similar to debris from
rocky bodies seen around white dwarfs \cite{Reach2009} would indicate a
rocky-body origin.  Compositions and sizes similar to the ISM would be
indicative of condensed dust in an M dwarf wind, similar to that formed
in the environments of more evolved stars \cite{Marigo2008}.

@@

It is currently unclear what, if any, relationship CPVs have to the
close-in rocky exoplanets that exist around most M dwarfs \cite{Dressing2015}.
However, 0.3-3\% of young M dwarfs show the CPV phenomenon
\cite{Rebull2020}, and our data show that the phenomenon occurs when
clumps of circumstellar material transit the star.  The implied
geometric correction suggests that an appreciable minority (10-30\%) of
young M dwarfs -- the rapidly rotating ones with centrifugal
magnetospheres -- host similar circumstellar environments to the CPVs.

@@

{\it Upper and Low Bounds on Dust}---An upper limit on the amount of
hot dust follows from the lack of an infrared excess.  A lower limit
follows if one assumes that most of the broadband optical depth comes
from dust absorption and scattering, rather than any radiative
processes associated with the plasma.

Three relevant objects for comparison to these limits include solar
prominences, planetesimals, and comets.
Prominences of the Sun typically have gas masses
of order $10^{14}$\,g-$10^{16}$\,g \cite{VialEngvold2015}.
A planetesimal of mass $\approx$10$^{15}$\,g with a bulk density of
1\,g\,cm$^{-3}$ would have a diameter of order one kilometer.
Halley's comet has a mass of order $10^{17}$\,g \cite{Rickman1989}, of
which $\sim$$10^{14}$\,g is shed per orbit, most of which inspirals
toward the Sun due to Poynting-Robertson drag.
Broadly, these calculations suggest that if dust is responsible for
the broadband variability of CPVs, it would need to be concentrated in
clumps with masses in the range of $10^{15}$-$10^{17}$\,g.

@@

\begin{figure}[!t]
  \centering
  \includegraphics[width=\textwidth]{figures/sf5.pdf}
  \caption{Time evolution of select regions in the Keck/HIRES
  spectra from 17 February 2025.
  The spectra have been smoothed with a Gaussian filter.
  The horizontal axis shows the relative velocity centered on each
  line's rest wavelength, normalized by the stellar equatorial
  velocity.  
  H$\alpha$ is the only Balmer line to show periodic variability of
  the form seen in Figure~\ref{fig:spec}.
  The He 5876\,\AA\ line shows a time-dynamic blueshift that 
  differs from the Balmer lines.
  Li 6708\,\AA\ shows no obvious absorption.
  }
  \label{fig:hirescuts}
\end{figure}

@@

\subsection{Modeling the Emitting Clump}
\label{subsec:model}

The physical conditions inside the circumstellar plasma clumps, in
particular the hydrogen number density, plasma temperature, ionization
fraction, and magnetic field strength can all be estimated using the
available data.
We caution that the following estimates are coarse: detailed
considerations of radiative transfer and detailed balance are a
worthwhile topic for future work, but are beyond our immediate scope.

Circumstellar H$\alpha$ emission might be sourced either from
scattering of stellar H$\alpha$ photons, or from recombination.  We
neglect scattering because we see the amplitude of the circumstellar
emission to vary by a factor of $\approx$5, which is not mirrored in
the chromospheric line core.  The volume emissivity under case B
recombination (in which Lyman-series photons are trapped by high
optical depth, so that recombinations cascade through H$\alpha$ and
other Balmer lines) can be written as
\begin{equation}
  j_{\rm H\alpha} = n_{\rm e} n_{\rm p} \alpha^{\rm eff}_{\rm H\alpha}(T) h \nu_{\rm H\alpha},
\end{equation}
where $n_{\rm e}$ and $n_{\rm p}$ are the electron and proton densities,
and $\alpha^{\rm eff}_{\rm H\alpha}(T)$ is the effective recombination
coefficient, i.e.\ the number of recombinations per unit volume, per
unit time.
For hydrogen with temperatures between 1,000 and 10,000\,K with
densities $n_{\rm H}$$\sim$10$^{10}$\,cm$^{-3}$,
$\alpha^{\rm eff}_{\rm H\alpha}$ is typically on the order of $10^{-12}$ to
$10^{-13}$\,cm$^3$\,s$^{-1}$ \cite{Hummer1987}.
Neglecting the effects of atoms other than hydrogen, we can assume
an ionization fraction $x$, such that $n_{\rm e} = n_{\rm p} = x
n_{\rm H}$, for $n_{\rm H}$ the hydrogen number density.  Let $L_{\rm
H\alpha} = j_{\rm H\alpha} V$, for $V$ the volume of the emitting
hydrogen.
The luminosity of circumstellar hydrogen emission, $L_{\rm H\alpha}$,
is an observable: our SED fitting routine yields
$L_\star$$\approx$0.012\,$L_\odot$, which gives
$L_{\rm H\alpha},\star$$\approx$1.0$\times$$10^{28}$erg\,s$^{-1}$, and
the luminosity of the clump $L_{\rm H\alpha}$ is of order one tenth
that of the star.
Assuming an emitting sphere of radius $r$, we can write
\begin{equation}
  n_H = \frac{1}{x}
  \left( 
    \frac{L_{\rm H\alpha}}{ \alpha^{\rm eff}_{\rm H\alpha}(T) h \nu_{\rm H\alpha} V }
  \right)^{1/2}.
\end{equation}
Taking nominal parameters of $r = 0.1 R_\star$, $x=1/2$, and
$\alpha^{\rm eff}_{\rm H\alpha}$=$10^{-13}$\,cm$^3$\,s$^{-1}$,
we find $n_{\rm H}\approx 5\times10^{11}$\,cm$^{-3}$.
Assuming a uniform density clump, this yields $M_{\rm H}\approx
8\times10^{16}$\,g.
A constraint on the magnetic field strength at the site of the clump
follows from the requirement that the magnetic pressure exceed the
thermal pressure, $B^2 / 8\pi > n_{\rm H} k T$.
Although we do not know the plasma temperature,
if it were significantly outside the range of
1,000 and 10,000\,K,
we would either full ionize the hydrogen, or not ionized enough of it.
We can therefore estimate
\begin{equation}
  B \gtrsim 3\,{\rm G}
  \left(
  \frac{n_{\rm H}}{5\times10^{11}\,{\rm cm}^{-3}}
  \frac{T}{5000\,{\rm K}}
  \right).
\end{equation}
Given that the average surface magnetic field strengths of low-mass
stars have been measured to span hundreds to thousands of Gauss
\cite{Kochukhov2021}, this bound is easily met.

>>>>

\begin{abstract}
\normalfont
Roughly one percent of red dwarfs younger than 100 million years
show structured, periodic optical light curves suggestive of
transiting opaque material that corotates with the star
\cite{Rebull2016,Stauffer2017,Rebull2018,Bouma2024}.  The
composition, origin, and even the existence of this material are
uncertain. The main alternative hypothesis is that these complex
periodic variables (CPVs) are explained by complex distributions of
bright or dark regions on the stellar surfaces
\cite{Koen2021}.  Here, we present time-series spectroscopy and
photometry of a rapidly-rotating ($P$=3.9\,hr)
CPV, TIC~141146667. The spectra show coherent sinusoidal Balmer
emission at twice to four times the star's equatorial velocity,
demonstrating the existence of extended clumps of corotating circumstellar
plasma --- a plasma torus.  Given that long-lived condensations of
cool ($10^4$ K) plasma can persist in the hot ($10^6$ K) coronae of
stars with a wide range of masses
\cite{CollierCameron1989,Townsend2005,Dunstone2006,Petit2013,Waugh2022,Daley-Yates2024},
these data support the idea that around the lowest-mass stars, such
condensations can be optically thick; the exact origin of the
optically thick material and its microphysical opacity remain unclear.
\end{abstract}

@@

The material's composition -- either pure plasma, or else a dusty plasma
-- can be clarified by time-series optical and infrared
spectrophotometry.  While observations of CPVs in the optical have
previously suggested chromaticity consistent with dust
\cite{Tanimoto2020,Gunther2022,Koen2023}, a gray opacity source such as
electron scattering in a plasma transiting over starspots might also
produce chromatic features \cite{Rackham2018}.  The composition and size
distribution of any dust that is present could be determined
by measuring the extinction curve for one or more CPVs from
1-10\,$\mu$m.  Composition and size distributions similar to debris from
rocky bodies seen around white dwarfs \cite{Reach2009} would indicate a
rocky-body origin.  Compositions and sizes similar to the interstellar
medium would be
indicative of condensed dust in an M dwarf wind, similar to that formed
in the environments of more evolved stars \cite{Marigo2008}.

@@

It is currently unclear what, if any, relationship CPVs have to the
close-in rocky exoplanets that exist around most M dwarfs
\cite{Dressing2015}.  However, 0.3-3\% of young M dwarfs show the CPV
phenomenon \cite{Rebull2020}, and our data show that the complex
photometric features occur when clumps of circumstellar material is
transiting the star.  The implied geometric correction suggests that
an appreciable minority (10-30\%) of young M dwarfs -- the rapidly
rotating ones with centrifugal magnetospheres -- host similar
circumstellar environments to the CPVs.

@@

@@

\begin{figure}[!t]
  \centering
  \includegraphics[width=\textwidth]{figures/sf5.pdf}
  \caption{Time evolution of select regions in the Keck/HIRES
  spectra from 17 February 2025.
  The spectra have been smoothed with a Gaussian filter, and cosmic
  rays have been removed.
  The horizontal axis shows the relative velocity centered on each
  line's rest wavelength, normalized by the stellar equatorial
  velocity.  
  H$\alpha$ is the only Balmer line to show periodic variability of
  the form seen in Figure~\ref{fig:spec}.
  The He 5876\,\AA\ line shows a time-dynamic blueshift that 
  differs from the Balmer lines.
  Li 6708\,\AA\ shows no obvious absorption.
  }
  \label{fig:hirescuts}
\end{figure}

@@

\subsection{Properties of the Plasma and Magnetospheric Environment}\phantom{+}
\label{subsec:gas}

The physical conditions inside the circumstellar plasma clumps, in
particular the hydrogen number density, plasma temperature, ionization
fraction, and magnetic field strength can all be estimated using the
available data.  We caution that the following estimates are coarse:
detailed considerations of radiative transfer and detailed balance are
a worthwhile topic for future work, but are beyond the scope of this
article.

Circumstellar H$\alpha$ emission might be sourced either from
scattering of stellar H$\alpha$ photons, or from recombination.  We
neglect scattering because we see the amplitude of the circumstellar
emission to vary by a factor of $\approx$5, which is not mirrored in
the chromospheric line core.  The volume emissivity under case B
recombination (in which Lyman-series photons are trapped by high
optical depth, so that recombinations cascade through H$\alpha$ and
other Balmer lines) can be written as
\begin{equation}
  j_{\rm H\alpha} = n_{\rm e} n_{\rm p} \alpha^{\rm eff}_{\rm H\alpha}(T) h \nu_{\rm H\alpha},
\end{equation}
where $n_{\rm e}$ and $n_{\rm p}$ are the electron and proton densities,
and $\alpha^{\rm eff}_{\rm H\alpha}(T)$ is the effective recombination
coefficient, i.e.\ the number of recombinations per unit volume, per
unit time.
For hydrogen with temperatures between 1,000 and 10,000\,K with
densities $n_{\rm H}$$\sim$10$^{10}$\,cm$^{-3}$,
$\alpha^{\rm eff}_{\rm H\alpha}$ is typically on the order of $10^{-12}$ to
$10^{-13}$\,cm$^3$\,s$^{-1}$ \cite{Hummer1987}.
Neglecting the effects of atoms other than hydrogen, we can assume
an ionization fraction $x$, such that $n_{\rm e} = n_{\rm p} = x
n_{\rm H}$, for $n_{\rm H}$ the hydrogen number density.  Let $L_{\rm
H\alpha} = j_{\rm H\alpha} V$, for $V$ the volume of the emitting
hydrogen.
The luminosity of circumstellar hydrogen emission, $L_{\rm H\alpha}$,
is an observable: our SED fitting routine yields
$L_\star$$\approx$0.012\,$L_\odot$, which gives
$L_{\rm H\alpha},\star$$\approx$1.0$\times$$10^{28}$erg\,s$^{-1}$, and
the luminosity of the clump $L_{\rm H\alpha}$ is of order one tenth
that of the star.
Assuming an emitting sphere of radius $r$, we can write
\begin{equation}
  n_H = \frac{1}{x}
  \left( 
    \frac{L_{\rm H\alpha}}{ \alpha^{\rm eff}_{\rm H\alpha}(T) h \nu_{\rm H\alpha} V }
  \right)^{1/2}.
\end{equation}
Taking nominal parameters of $r = 0.1 R_\star$, $x=1/2$, and
$\alpha^{\rm eff}_{\rm H\alpha}$=$10^{-13}$\,cm$^3$\,s$^{-1}$,
we find $n_{\rm H}\approx 5\times10^{11}$\,cm$^{-3}$.
Assuming a uniform density clump, this yields $M_{\rm gas}\approx
8\times10^{16}$\,g.

Finally, a constraint on the magnetic field strength at the site of the clump
follows from the requirement that the magnetic pressure exceed the
thermal pressure, $B^2 / 8\pi > n_{\rm H} k T$.
Although we do not know the plasma temperature,
if it were significantly outside the range of
1,000 and 10,000\,K,
we would either full ionize the hydrogen, or not ionized enough of it.
We can therefore estimate
\begin{equation}
  B \gtrsim 3\,{\rm G}
  \left(
  \frac{n_{\rm H}}{5\times10^{11}\,{\rm cm}^{-3}}
  \frac{T}{5000\,{\rm K}}
  \right).
\end{equation}
Given that the average surface magnetic field strengths of low-mass
stars have been measured to span hundreds to thousands of Gauss
\cite{Kochukhov2021}, this bound is easily met.

\subsection{Upper and Low Bounds on Dust}\phantom{+}
\label{subsec:dust}

The material's exact composition -- either pure plasma, or else a
dusty plasma -- is not known.  In particular, the idea of dust being
present seems plausible given observations of chromatic transits in
analogous objects \cite{Tanimoto2020,Gunther2022,Koen2023}.
This scenario is highly constrained.  An upper limit on the amount of
hot dust follows from the lack of an infrared excess.  A lower limit
follows if one assumes that most of the broadband optical depth comes
from dust absorption and scattering, rather than any radiative
processes associated with the plasma.

Three relevant objects for comparison to these limits include solar
prominences, planetesimals, and comets.  Prominences of the Sun
typically have gas masses of order $10^{14}$\,g-$10^{16}$\,g
\cite{VialEngvold2015}.  A planetesimal of mass $\approx$10$^{15}$\,g
with a bulk density of 1\,g\,cm$^{-3}$ would have a diameter of order
one kilometer.  Halley's comet has a mass of order $10^{17}$\,g
\cite{Rickman1989}, of which $\sim$$10^{14}$\,g is shed per orbit,
most of which inspirals toward the Sun due to Poynting-Robertson drag.

Broadly, our calculations suggest that if dust is responsible for the
broadband variability of CPVs, it would need to be concentrated in
clumps with masses in the range of $10^{15}$-$10^{17}$\,g.  Given
$M_{\rm gas}$$\approx$8$\times$10$^{16}$\,g from
Section~\ref{subsec:gas}, this range spans mass ratio ranges $M_{\rm
gas}/M_{\rm dust}$ of 100 to 1.  More careful measurements of this
ratio---in particular enabled by inferring the dust mass through high
precision infrared spectrophotometry---would provide a pathway for
distinguishing between environments that more closely resemble the
interstellar medium versus protoplanetary disks.
